---
title: "Graphics systems in R"
output: word_document
---

```{r setup, cache=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width=4, fig.height=4, dpi=100, warnings=FALSE, message=FALSE)
```
``` {r include = FALSE }
oldPar = par()
```

## III. Grammar of Graphics

{ggplot2} is a high-level R package developed by Hadley Wickham based on Leland Wilkinson's concept of the grammar of graphics ("The Grammar of Graphics", 1999/2005).

In practical sense, with ggplot2 building blocks of a plot are independently specified and then combined to create complex multi-layered graphics.

Main building blocks  

- data
- aesthetic mapping 
- geometric object 
- statistical transformations
- coordinate system
- scales
- themes
- faceting

Basic plotting functions  

- qplot() - makes quick plots, resembles standard graphics
- ggplot() - the main plot function, allows detailed control of the graphical elements
 
##### --------------------------------------------------------------------------------------------------------

### 1. Set Environment
```{r message = FALSE, warning = FALSE}
setwd("/Users/balaji/Downloads/plots")
load('helperFunctions.RData')
library(ggplot2)
library(reshape2)
library(grid)
library(gridExtra)
```

### 2. Set Data
``` {r }
load('/Users/balaji/Downloads/plots/NeveG.RData')
Data <- NeveG_data # data.frame, wide form
mData <- melt(Data, measure.vars = 5:13, variable.name = 'Gene', value.name = 'GEX') # data.frame, long form
```

### 3. quickplot()
```{r fig.width = 6, fig.height = 6}
# prepare histograms
f1A <- qplot(data = Data, ageY , geom=c('histogram'))
f1B <- qplot(data = Data, MEGF9, geom=c('histogram'))
# prepare plot with data points and smoothed line
f2 <- qplot(data = Data, ageY, MEGF9, geom=c('point','smooth'))
# prepare boxplots
f3  <- qplot(data = Data, ER, ageY, geom = c('boxplot'))
# arrange plots
grid.arrange(f1A, f1B, f2, f3, ncol=2)
```

### 4. ggplot()

- 4.1 mapping data to aesthetics (choosing axis, color/fill, shape/linetype, size)
``` {r }
plot.data <- ggplot( data = mData, aes(x = geneCluster, y = GEX, fill = geneCluster) )
plot.blanc <- ggplot( data = mData, aes(x = geneCluster, y = GEX, color = geneCluster) )

```

- 4.2 geoms (the actual marks plotted)
``` {r }
violin <- geom_violin()
dot <- geom_jitter(shape = 16, position = position_jitter(0.2), size=1)
box <- geom_boxplot(outlier.size = -1, width = 0.25, fill = 'antiquewhite', color = 'antiquewhite', alpha = 0.5)
```

- 4.3 stat 
``` {r }
stat.median <- stat_summary(fun.data=median, geom="point", shape=23, size=2)
stat.error <- stat_summary(fun.data = mean_err, geom="pointrange")
```

- 4.4 theme and plot attributes
``` {r }
theme <- theme_bw()
leg <- theme(legend.position = 'none')
titles <- labs(title = "Violin Plots", x = "", y = "log2 gene expression")
```

- 4.5 scales
``` {r }
paint.fill <- scale_fill_manual(values = c("red","orange", "blue"))
paint.col <- scale_color_manual(values = c("red","orange", "blue"))
y.axis <- scale_y_continuous(breaks = seq(2,14,2))
```

- 4.6 combine layers
``` {r }
plot.data + violin + dot + box + titles + leg + y.axis + paint.fill + coord_flip()
plot.blanc + geom_violin(trim = FALSE) + stat.error + titles + theme + leg + paint.col
```

### 5. Faceting

- 5.1 Define subsets as the levels of a single grouping variable with facet_wrap()
``` {r }
dat.plot <- ggplot(data = mData, aes(x = GEX, fill = ER))
histo <- geom_histogram(binwidth=1, position = 'identity', alpha = 0.6)
panel.wrap <- facet_wrap(~Gene)

dat.plot + histo + panel.wrap + labs(x = 'log2 Expression') + theme_light()
```

- 5.2 Define subsets as the crossing of two grouping variables with facet.grid()
```{r }
dat.plot <- ggplot(data = mData, aes(x = GEX, fill = ER))
histo <- geom_histogram(binwidth=1, position = 'identity', alpha = 0.6)
panel.grid <- facet_grid(ER ~ geneCluster)
theme.bg <-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

dat.plot + histo + panel.grid + theme.bg

```

### 5. Good ggplot2 summary from R Studio

[ggplot2-cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

### 6. Exercises

- 6.1 inspect R dataset: diamonds

```{r results = 'hide'}
library(tidyverse)
data(diamonds)
diamonds
?diamonds
summary(diamonds)
```
- 6.2 make plots:
1. histogram + density plot for the "price" variable"; hint: use geom_histogram(aes(y=..density..)) and geom_density
2. scatterolot (carat-price) + non-linear smoother by cut; color points by cut

```{r }
ggplot(diamonds, aes(x = price)) + 
  geom_histogram(aes(y=..density..), fill='lightblue', col='black') + 
  geom_density(col='blue', fill='yellow', alpha=0.1)

ggplot(diamonds, aes(x=carat, y=price,color=cut)) +
  geom_point(alpha=0.5, size=0.5) + geom_smooth()
```

### R Session Info
```{r}
sessionInfo()
```
