---
title: "Graphics Systems in R"
output: word_document
---

```{r setup, cache=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width = 3.6, fig.height = 3.6, dpi = 100)
```
``` {r include = FALSE}
oldPar = par()
```

## II. Grid graphics

{grid} is an alternative, low-level graphics system in R developed by Paul Murrell in the 1990's. {grid} provides large flexibility in the layout of graphics and enables creating of editable graphical objects. {grid} itself does not provide high-level functions to create complete plots, but is a graphics engine behind other packages such as {lattice} and {ggplot2}.

- an add-on package that needs to be loaded into an R session
- viewport defines context for drawing (location and size)
- grob (graphical object) defines drawing shape, e.g., line, circle, polygon, etc.
- gpar(parameter = value) sets parameters of the graphics
- get.gpar() lists the gpar parameters

***


### 1. Set Environment  
```{r }
setwd("/Users/balaji/Downloads/plots")
load('helperFunctions.RData')
library('grid')
```

### 2. Create Viewports and Plot Shapes  

vieport() defines a rectangular region and its coordinates; multiple viewports can be defined within the plotting area. A default plotting region is defined with "npc" units (normalised parent coordinates), where the origin of the viewport is (0, 0) and the viewport has a width and height of 1 unit;  for example, (0.5, 0.5) is the centre of the viewport,  where left corner is (x=0, y=0) and right corner is (x=1, y=1). 

```{r }
# draw borders of the starting viewport:
grid.rect(gp = gpar(lty='dashed'))
grid.text(label = "(0,0)", x = 0+0.05, y=0+0.03)
grid.text(label = "(1,1)", x = 1-0.05, y=1-0.03 )

# define viewports:
VP1 <- viewport(x = 0.25, y = 0.75, width = 0.4, height = 0.4)
VP2 <- viewport(x = 0.75, y = 0.25, width = 0.4, height = 0.4)
VP3 <- viewport(x = 0.5, y = 0.5, width = 0.4, height = 0.4, angle = 45)

# open viewports and draw graphical elements:
#...square, x-axis, points
pushViewport(VP1)
  grid.rect()
  grid.xaxis( at = seq(0, 1, by = 0.2))
  grid.circle(x = c(0.2, 0.6, 0.8), y = c(0.4, 0.8, 0.5), r=0.02)
upViewport()

#...circle in a box
pushViewport(VP2)
  grid.rect(gp = gpar(fill = 'black'))
  grid.circle()
upViewport()

#...rotate green box / repeat a few times to see the effect in child viewports
pushViewport(VP3)
  grid.rect( gp = gpar(fill = 'lightgreen') )

  upViewport()
  
#..read and insert a png image
gridImage(image.file = '/Users/balaji/Downloads/plots/Hela.png')

#..nested viewports
for(i in 1:10)
{
  pushViewport(VP3)
  grid.rect( gp = gpar(fill = 'lightgreen') )
}
upViewport(0)
```

### 3. Generate grobs

- draw and edit shapes with grid.*()

``` {r fig.width=2, fig.height=2}
# draw
pushViewport(VP1)
grid.rect(name="box")
grid.circle(r=0.1, gp=gpar(fill='black'))

# edit and change a parameter
grid.edit("box", gp=gpar(fill="lightyellow3"))

# remove
grid.remove("box")
#upViewport()
```

### 4. Combine grobs
- store and edit shapes with \*Grob(); \*Grob produces a graphical object, but no output
- list objects with gList() and draw with grid.draw() 

``` {r fig.width = 3, fig.height = 3}
spot <- circleGrob(r=0.3)
spot <- editGrob(spot, gp = gpar(fill = 'blue'))
txt <- textGrob('Hello world!', gp = gpar(col = 'white', cex = 2))
my.grob <- gList(spot, txt)
grid.draw(my.grob)
```

#### if interested in more try  [https://www.stat.auckland.ac.nz/~paul/grid/grid.html](https://www.stat.auckland.ac.nz/~paul/grid/grid.html)

### 5. {lattice} and Trellis Plots 

graphs that help examine multivariate relationships by plotting a variable conditioned on one or more other variables.

### 5. Load lattice and data
``` {r }
library(lattice)
load('/Users/balaji/Downloads/plots/NeveG.RData')
Data <- NeveG_data
```

- 5.1 density plots
``` {r fig.width = 8, fig.height = 8}
densityplot(~ABAT+CALD1+ESR1|ER*geneCluster, data=Data,
            main = 'Density plots by ER status and gene cluster', 
            auto.key=list(space='right'))

```

- 5.2 scatter plots
``` {r fig.width = 5, fig.height = 8}

# group x,y and condition x-y relationship on one variable
splom(~Data[,5:7]|Data$ER, groups = Data$geneCluster,
      layout = c(1,2),
      main = 'Scatter plots by ER status', 
      auto.key = list(space='top', title='Cluster'))
```

- 5.3 box plots with conditioning on a continous variable
``` {r fig.width = 8, fig.height = 4}
# cut continuous variable into intervals
age <- cut(Data$age, 3)
b <- bwplot(~Data$LARGE|age*Data$ER, main = 'LARGE expression by ER status and age category', xlab = 'log2', ylab = 'Box plots')
update(b, fill = 'lightgrey')
```

- 5.4 See R help pages for more types of lattice plots
``` {r echo=FALSE}
#help(lattice)
```
- 5.5 Exercise: finish the combined grid/lattice plot:  
1. create two viewports (blue and grey here); hint: use npc coordinates (0 to 1) and center of colored rectangles;
2. use print(scatterLattice, newpage=FALSE) to place the lattice plot in the grey viewport

```{r }
grid.rect(gp=gpar(fill='black', lty='solid'))
VP1 <- viewport(x = 0, y = 0.5, width = 0.25, height = 1)
pushViewport(VP1)
grid.rect(gp=gpar(fill="steelblue3"))
grid.text("Figure1", x=0.7,y=0.98,gp=gpar(cex=1.5, col='black'))
upViewport()
grid.text("MEGF9 ~ Age|Cluster", x=0.5, y=0.95, gp=gpar(cex=2,
                                    col='steelblue3'))
VP2 <- viewport(x = 0.6, y = 0.5,width = 0.6,height = 0.6)
pushViewport(VP2)
grid.rect(gp=gpar(fill='lightgrey', lty='solid'))

scatterLattice <- xyplot(MEGF9 ~ ageY|geneCluster, data=Data, layout=c(3,1),
                         type = c("p","r"), pch=16, col='black')
print(scatterLattice, newpage=FALSE)
```

```{r echo = FALSE, fig.width=6, fig.height=6}
grid.rect(gp=gpar(fill='black', lty='solid'))
scatterLattice <- xyplot(MEGF9 ~ ageY|geneCluster, data=Data, layout=c(3,1), type = c("p","r"), pch=16, col='black')
grid.text("MEGF9 ~ Age|Cluster", x=0.5, y=0.95, gp=gpar(cex=2, col='steelblue3'))
VP1 <- viewport(x = 0.1, y = 0.5, width = 0.2, height = 1)
VP2 <- viewport(x = 0.6, y = 0.5, width = 0.6, height = 0.6)
pushViewport(VP1)
grid.rect(gp=gpar(fill='steelblue3', lty='solid'))
grid.text("Figure 1", x=0.5, y=0.95, gp=gpar(cex=2))
upViewport()
pushViewport(VP2)
grid.rect(gp=gpar(fill='grey', lty='solid'))
print(scatterLattice, newpage=FALSE)
upViewport()
```


### R Session Info
``` {r}
sessionInfo()
```
