---
title: "Graphics systems in R"
output: word_document
---

```{r setup, cache=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width=4, fig.height=4, out.height = '30%', out.width = '30%', dpi=100)
```
``` {r include = FALSE}
oldPar = par()
```

## I. Standard graphics

Base package {graphics} is the graphics system that was originally developed for S language at Bell Labs in the 1970s.  It is loaded by default when R starts.

- plot() is the main, high-level function for creating graphics
- par() sets-up parameters which control the layout and look of the graphics  

##### --------------------------------------------------------------------------------------------------------

### 1. Set Environment
```{r }
setwd("/Users/balaji/Downloads/plots")
load('helperFunctions.RData')
```

### 2. Set Plotting data

Generate random data points ("Variable" - R vector) from standard normal distribution (N=50) 

```{r }
set.seed(100); Variable <- rnorm(50)
set.seed(100); names(Variable) <- randomNames(n=50, length=3)
```

### 3. Generic X-Y Plot
```{r }
plot(Variable) ## type = 'p'; points
```

- Add figure regions in the background
```{r }
plotRegions() 
```

### 4. Customize Plot  

Graphical options can be customized through:

- an argument in par() function; par( parameter.name = value ); the option will be in effect until the end of the R session or until updated
- an argument in a plot() function; plot(..., parameter.name = value); the option will be in effect for the current plot
- current values of the parameters can be viewed with par() and par()$parameter.name or par("parameter.name")

Examples:  

``` {r results = 'hide'}
# list values for all parameters 
# par()
# ?par
length(par())

# figure/plot dimensions in inches
par()$fin; par()$pin

# inner/outer figure margins in inches and the number of text lines
par()$mai; par()$mar
par()$omi; par()$oma

# size of plotting text and symbols (default = 1; 1.5 is 50% larger, 0.5 is 50% smaller, etc.
par()$cex;par()$cex.axis; par()$cex.label; par()$cex.main

# shape of plotting symbol
par()$pch

# line width and type
par()$lwd; par()$lty
```


- 4.1 Modify margins and add annotation in the outer margin area; add figure and outer boxes
``` {r }
par( oma = c(0,0,2,0), mar = c(6,4,2,1) )
plot(Variable)

mtext( 'Page 1', side = 3, out = TRUE, cex = 0.5 )
box(which = 'outer')
box(which = 'figure')
```

- 4.2 Add/modify titles
``` {r }
par( oma = c(0,0,2,0), mar = c(6,4,2,1) )
plot(Variable,
     
     main = 'Figure 1', cex.main = 2, 
     sub = 'sample size = 50', cex.sub = 0.8,
     xlab = 'Subject', ylab = 'Z-score')

mtext( 'Page 1', side = 3, out = TRUE, cex = 0.5 )
box(which = 'outer')
box(which = 'figure')
```

- 4.3 Plot standard point shapes
``` {r fig.width = 3, fig.height = 3, out.width = '50%', out.height = '50%' }
plotPoints() 
```

- Change point shape, color, and size
``` {r }
par( oma = c(0,0,2,0), mar = c(6,4,2,1) )
plot(Variable,
     main = 'Figure 1', cex.main = 2, 
     sub = 'sample size = 50', cex.sub = 0.8,
     xlab = 'Subject', ylab = 'Z-score',
     
     type = 'p', pch = 16, col = 'orange', cex = 1.5)

mtext( 'Page 1', side = 3, out = TRUE, cex = 0.5 )
box(which = 'outer')
box(which = 'figure')
```

- 4.4 Draw customized axes
``` {r }
par( oma = c(0,0,2,0), mar = c(6,4,2,1) )

# suppress drawing of both axes, modify limits for y-axis

plot(Variable,
     main = 'Figure 1', cex.main = 2, 
     sub = 'sample size = 50', cex.sub = 0.8,
     xlab = 'Subject', ylab = 'Z-score',
     pch = 16, col = 'orange', cex = 1.5,
     
     axes = FALSE,
     ylim = c(-3,3))

# add axes

axis(1, at = seq(0, 50, 5))
axis(2, at = seq(-3, 3, by=0.5), labels=FALSE, tck=-0.01)
axis(2, at = seq(-3, 3, by=1), las=2)
box(which = 'plot', bty= 'o')

mtext( 'Page 1', side = 3, out = TRUE, cex = 0.5 )
box(which = 'outer')
box(which = 'figure')
```
  
- 4.5 Plot line types
``` {r fig.width = 3, fig.height = 2, out.width = '50%', out.height = '30%'}
plotLines()
```

- 4.6 Add horizontal line and label maximal point
``` {r }
par( oma = c(0,0,2,0), mar = c(6,4,2,1) )

# suppress drawing of both axes, modify limits for y-axis

plot(Variable,
     main = 'Figure 1', cex.main = 2, 
     sub = 'sample size = 50', cex.sub = 0.8,
     xlab = 'Subject', ylab = 'Z-score',
     type = 'p', pch = 16, col = 'orange', cex = 1.5,
     axes = FALSE,
     ylim = c(-3,3))

# add axes

axis(1, at = seq(0, 50, 5))
axis(2, at = seq(-3, 3, by=0.5), labels=FALSE, tck=-0.01)
axis(2, at = seq(-3, 3, by=1), las=2)
box(which = 'plot', bty = 'o')
mtext( 'Page 1', side = 3, out = TRUE, cex = 0.5 )
box(which = 'outer')
box(which = 'figure')

abline( h = 0, lty = 'longdash' )

maxValue <- max(Variable)
index <- which( Variable == maxValue )
label <- names(Variable)[index]
text(x = index, y = maxValue+0.4, labels = label)
```
  
- 4.7 Exercise - make a plot using "Variable" object:
1. width = 3 in, height = 3 in;
2. all four outer margins = 3 lines;
3. plot data points labels in place of data points;
   hint: plot(1:10, type = 'none') will plot figure frame, but not data points;
4. omit plotting x-axis and its label;
5. modify y-axis label, add figure title, and add date in the bottom outer margin;
6. plot only first character of data label
   hint: use substring(string, first, last) to extract the first character;
7. plot data labels one half size;
8. use color = 1:7 to color the text;
9. add figure box.

```{r echo = FALSE, fig.width = 5, fig.height = 5}
par( pin = c(3,3), oma = c(3,3,3,3) )
plot(Variable,type = "n",xaxt="n",xlab="",ylab="ylab",axes = F)
title("main")
axis(2, at = seq(-2, 2, by=1), labels=T)
mtext( Sys.Date(), side =1, out = TRUE )
labels = substring(names(Variable),1,1)
text(x = c(1:length(Variable)), y = Variable,labels = labels,col =1:7 )
box(which = 'figure')
```


``` {r include = FALSE}
options(warn = -1); par(oldPar); options(warn = 0)
```

## 5. Example Dataset  

##### NeveRM, "A collection of breast cancer cell lines for the study of functionally distinct cancer subtypes"; [PMC2730521](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2730521/)

- 5.1 load data
``` {r }
load("/Users/balaji/Downloads/plots/NeveG.RData")
Data <- NeveG_data
```
- 5.2 look at the data
``` {r results = 'hide'}
head(Data)
str(Data)
summary(Data)
```

- 6.3 calculate descriptive statistics
``` {r }
temp = Data[,-c(1:4)]
means <- apply(temp, 2, mean, na.rm = T)
medians <- apply(temp, 2, median, na.rm = T)
stdevs <- apply(temp, 2, sd, na.rm = T)
sems <- apply(temp, 2, function(x){
              n <- sum(!is.na(x))
              sd(x, na.rm = T)/sqrt(n)})
size <- apply(temp, 2, function(x) sum(!is.na(x)))
cv <- stdevs/means*100

## combine stats with data.frame() function
Stat <- data.frame(N = size, Mean = means, Median = medians, SD = stdevs, SEM = sems, CV = cv)

## save as *.RData object
save(Stat, file = 'Stats.RData')

## save as .csv file
write.csv(Stat, file = 'Stats.csv')
```

## 6. Generate Plots

### 6.1 Histogram
``` {r fig.width = 12, fig.height = 4, out.width = '60%', out.height = '60%'}
par(mfrow = c(1,2))

hist(Variable) ## freq = TRUE/FALSE; breaks = the number of bins

# overlay normal curve
hist(Variable, freq = FALSE, col = 'skyblue', border = 'black')
curve(dnorm(x, mean = mean(Variable), sd = sd(Variable)), lwd=2, add=T)
```

### 6.2 Density Plot
``` {r fig.width = 12, fig.height = 4, out.width = '60%', out.height = '60%'}
par(mfrow = c(1,2))
Density = density(Variable) ## kernel density estimate (smoothing function)
plot(Density)
plot(Density, main = 'Density plot', sub = '', xlab = 'log2', lwd = 2)
lines(density(Variable+0.5), lwd = 2, col = 'red')
```

### 6.3 Dot Plot
``` {r fig.width = 10, fig.height = 5, out.width = '60%', out.height = '60%'}
par(mfrow = c(1,2))
stripchart( Data[, c('ABAT', 'CALD1')],
    vertical = TRUE, method = 'stack',
    pch = 16, col = 'green3',
    ylab = 'log2 Expression', main = 'Dot plot')
x.coord = c(1,2)
y.coord = Stat[c('ABAT','CALD1'), 'Median']
segments(x0 = x.coord+0.2,  y0 = y.coord, x1 = x.coord-0.2, y1 = y.coord,
  lwd = 2, cex = 1.5)

# use formula to plot data points by category
stripchart( ESR1 ~ geneCluster, data = Data,
    vertical = FALSE, method = 'jitter', jitter=0.2, 
    pch = 16, col = c('red', 'orange', 'blue'), 
    group.names = levels(Data$geneCluster),
    main = 'Dot plot',
    las =2)
```

### 6.4 Box Plot
``` {r fig.width = 8, fig.height = 4, out.width = '50%', out.height = '50%'}
boxplot(Data[, 5:10], col = 2:7, main='Box plot', ylab = 'log2', boxwex = 0.3)
# boxplot(Data[, 5], plot=F)
```

### 6.5 Combine Box plot and Dot plot
``` {r  fig.width = 10, fig.height = 5, out.width = '50%', out.height = '50%'}
library(reshape2)
mData <- melt(Data, measure.vars = 5:13, variable.name = 'Gene', value.name = 'GEX')

# combine box plot and  dot plot
par(oma = c(2,0,0,0))
boxplot(GEX ~ ER+Gene, data = mData,
        outline = FALSE, varwidth = TRUE,
        col = c('coral','lightblue'),
        main='Box plot + Dot plot',ylab = 'log2',
        las = 2)
stripchart(GEX ~ ER+Gene, data = mData,
           method = 'jitter', vertical = TRUE,
           pch=16, add=T)

```

### 6.5 Scatter Plot and Correlation
``` {r  fig.width = 4, fig.height = 4, out.width = '20%', out.height = '20%'}

# set variables
x <- Data$ABAT
y <- Data$ESR1
limit <- range(c(x,y))
r <- round(cor(x,y),2)

# make plot
par(pty='s')
plot(x, y, ylim = limit, xlim = limit, xlab = 'ABAT', ylab = 'ESR1')
abline(0,1)
fit <- lm(y ~ x)
abline(fit, col='blue')
legend('bottomright', col = c('black','blue'), lty = 1, lwd = 2, 
       legend = c('identity line', 'regression line'), cex=0.7)
mtext(paste('r', r, sep=' = '))
title('Scatter plot and correlation')
```

``` {r  fig.width = 6, fig.height = 4, out.width = '20%', out.height = '20%'}
# R object built-in diagnostic graphics
par(mfrow = c(2,2), pty='m', mar = par()$mar+c(-1,0,-2,0))
plot(fit)
```

### 6.5 Bar Plots

``` {r  fig.width = 6, fig.height = 4, out.width = '20%', out.height = '20%'}
par(mfrow = c(1,2))

# plot ordered means
d <- Stat$Mean
name <- rownames(Stat)
ord <- order(d)
barplot( 2^d[ord], names = name[ord], las=2, density = 20, col='lightblue', border='darkblue')
title(ylab = 'Mean Expression', main = 'Bar plot')

# calculate group estimates (means, standard errors), t- and p-value
x <- Data$TFF1[Data$ER == '[+]']
y <- Data$TFF1[Data$ER == '[-]']
name <- c('ER+','ER-')
x.mean <- mean(x)
y.mean <- mean(y)
x.sem <- sd(x)/sqrt(length(x))
y.sem <- sd(y)/sqrt(length(y))

test <- t.test(x=x, y=y, var.equal = TRUE)
tval <- round(test$statistic, 2)
pval <- signif(test$p.value, 1)


# plot barplot with error bars
d <- data.frame(Mean = c(x.mean, y.mean), SEM = c(x.sem, y.sem))
yaxis.max <- max(d$Mean + d$SEM)
bar <- barplot( d$Mean, names = name, ylab = 'log2', ylim = c(0, yaxis.max))
segments(x0 = bar, y0 = d$Mean-d$SEM, x1 = bar, y1 = d$Mean+d$SEM, lwd=2)
title(sub = paste('P-value = ', pval, ', t-value = ', tval, sep = ''), 
      main = 'Bar plot with error bars')


# plot barplot with error bars
d <- data.frame(Mean = c(x.mean, y.mean), SEM = c(x.sem, y.sem))
yaxis.max <- max(d$Mean + d$SEM)
bar <- barplot( d$Mean, names = name, ylab = 'log2', ylim = c(0, yaxis.max))
segments(x0 = bar, y0 = d$Mean-d$SEM, x1 = bar, y1 = d$Mean+d$SEM, lwd=2)
title(sub = paste('P-value = ', pval, ', t-value = ', tval, sep = ''), 
      main = 'Bar plot with error bars')
```

- 6.5.1 Exercise - plot error bars using arrows() function:  
hint: check ?arrows arguments;  
modify 'angle', 'code' (necessary), and 'length' (optional)
```{r echo=FALSE}
d <- data.frame(Mean = c(x.mean, y.mean), SEM = c(x.sem, y.sem))
yaxis.max <- max(d$Mean + d$SEM)
bar <- barplot( d$Mean, names = name, ylab = 'log2', ylim = c(0, yaxis.max))
arrows(x0 = bar, y0 = d$Mean-d$SEM, x1 = bar, y1 = d$Mean+d$SEM, lwd=2,
       code=3,length = 0,angle = 30)
title(sub = paste('P-value = ', pval, ', t-value = ', tval, sep = ''), 
      main = 'Bar plot with error bars')
```

### R Session Info
``` {r}
sessionInfo()
```
