---
title: "Working with Maps"
output: word_document
always_allow_html: yes
---

```{r setup, cache=FALSE, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width=4, fig.height=4, dpi=300, warning=FALSE, message=FALSE)
```
``` {r include = FALSE }
oldPar = par()
```

## I. Package {leaflet}

Leaflet is one of the most popular open-source JavaScript libraries for interactive maps. This R package makes it easy to integrate and control Leaflet maps in R.

##### --------------------------------------------------------------------------------------------------------

### 1. Set Environment  

``` {r  warning = FALSE}
setwd("C:/Users/YO/Documents/R Graphics")

library(datasets); library(leaflet); library(maps)
library(grid); library(gridExtra); library(OpenStreetMap); library(rgdal)
```

### 2. The Map Widget  

- 2.1 Basic rules  

leaflet() returns a map widget object, to which graphics elements and layers can be added up with the "%>%" operator.  

``` {r results = 'hide'}

df = data.frame(Lat = -15:15, Long = 15:-15)

m <- leaflet(df) %>% 
  
  addCircles(~Lat, ~Long, col = 'red') %>%
  addRectangles(lat1 = 0, lng1 = 0, lat2 = 10, lng2 = 10, fill=F) %>%
  addMarkers(lat = c(7,4), lng = c(7,-4), popup = 'text') %>%
  addCircleMarkers(lat = 5, lng = 5, radius = 20, color = 'orange', fill = TRUE) %>%
  setView(lat = 5, lng = 5, zoom = 0) %>% addTiles() %>%
  setView(lat = 5, lng = 5, zoom = 4) %>%
  setView(lat = 5, lng = 5, zoom = 19) %>%
  setView(lat = 5, lng = 5, zoom = 5) %>%
  addPopups(lat = 10, lng = 4, "square", options = popupOptions(closeButton = FALSE))

print(m)

?addCircles
```
- 2.1 An example

[coordinates can be found here]:(http://www.distancesfrom.com/Latitude-Longitude.aspx)

``` {r }
Lat = 39.0025
Long = -77.109
txt = '<b>FAES</b></p>Nov 16-17'

# Create map object
m <- leaflet() %>%

# Add default OpenStreetMap map tiles
addTiles() %>%

# Zoom at the location
setView(lat = Lat, lng = Long, zoom = 16) %>%

# Add popup
addMarkers(lat = Lat, lng = Long, popup = txt) %>%

# Add filled marker
addCircleMarkers(lat = Lat, lng = Long, radius = 30, color = 'cornflowerblue', fill = TRUE)

print(m)

# Export map from RStudio browser to png or html file
```

### 2. Use data with maps

- 2.1 Use paths to add polygons
``` {r }
library(maps)
mapStates = map("state", fill = TRUE, plot = FALSE)
leaflet(data = mapStates) %>% addTiles() %>%
  addPolygons(fillColor = topo.colors(10, alpha = NULL), stroke = FALSE)
```


- 2.2 Make choropleth map  

[data from]:(https://github.com/bahar/WorldCityLocations):  visulize citie's altitude

``` {r }
#https://github.com/bahar/WorldCityLocations
wct <- read.delim("misc/World_Cities_Location_table.txt", stringsAsFactors=FALSE, sep=';',head=F) 
colnames(wct) <- c('id', 'country', 'city', 'latitude', 'longitude', 'altitude')


pal <- colorQuantile("YlOrRd", NULL, n = 9)

city_popup <- paste0("<strong>State: </strong>", 
                      wct$country, 
                      "<br><strong>City: </strong>", 
                      wct$city,
                      "<br><strong>altitude: </strong>", 
                      wct$altitude)

m <- leaflet(data = wct) %>% addTiles() %>%
        addCircleMarkers(fillColor = ~pal(wct$altitude),
          radius = 2, stroke = FALSE, fillOpacity = 1,
           popup = city_popup) %>%
            setView(0,0, zoom = 2)
print(m)             
```
