---
title: "Graphics device"
runtime: html
output:
  word_document: default
  html_document:
    keep_md: yes
---

```{r setup, cache=FALSE, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width=4, fig.height=4, dpi=300, warning=FALSE, message=FALSE)
```
``` {r include = FALSE }
oldPar = par()
```

## I. Interactive Displays

##### --------------------------------------------------------------------------------------------------------

### 1. Set Environment  

``` {r  warning = FALSE}
setwd("/Users/balaji/Downloads/plots")

# load data
load('/Users/balaji/Downloads/plots/NeveG.RData')
load('/Users/balaji/Downloads/plots/NeveG200.RData')

library(ggvis); library(DiagrammeR); library(dygraphs); library(leaflet); library(reshape2);
library(graph); library(Rgraphviz); library(ggplot2); library(datasets); library(forecast); library(ggfortify); library(grid); library(gridExtra); library(OpenStreetMap); library(rgdal)
library(RColorBrewer); library(colorspace); library(colorRamps); library(dendextend)
library(gplots); library(hexbin); library(plotly); library(heatmaply)
```

### 2. Set data  
``` {r  warning = FALSE}
Data <- NeveG_data
mData <- melt(Data, measure.vars = 5:13, variable.name = 'Gene', value.name = 'GEX')

```

### 3. [{ggvis}](http://ggvis.rstudio.com/)  

#### An implementation of an interactive grammar of graphics similar to ggplot2. 

- the "%>%"" operator layers up graphics elements; ("+" in ggplot2)  
- initial object is created with ggvis(); (ggplot in ggplot2)    
- actual plots are made with layers; (geoms in ggplot2)    
- "~" is always placed before variable name in a plotting formula    
- "=" maps a variable to a property, e.g fill = ~Var would set the color based on the levels of Var  
- ":=" sets a property based on a value, e.g. fill := 'blue' would set the color to blue  
- interactive inputs are introduced with properties or input_* functions   

#### 3.1 Histogram
```{r fig.width = 3.5, fig.height = 1.5}

Data %>%
  ggvis(~MEGF9) %>%
  
  # change color on hover
  layer_histograms(fill.hover := 'white', 
        
        # change bin width
        width = input_slider(0.1, 2, step = 0.1, label = "Bin Width"),
        
        # choose bar color
        fill := input_select(choices = c('aliceblue', 'lightyellow', 'lavender','lightgrey'),                             selected = "aliceblue", label = "Fill Color")) %>%
  add_axis("x", title = "MEGF9 log2 Expression") %>%
  add_axis("y", title = "Bin Count")
```

#### 3.2 Scatter plot
```{r fig.width = 3.5, fig.height = 1.5}

Data %>%
  ggvis(~LARGE, ~ESR1,
        
        # change point size
        size := input_numeric(50, label = "Point size"),
        
        fill = ~ER,
        
        # change opacity
        opacity := input_slider(0, 1, value = 1, label = "Opacity")) %>% 
  
  layer_points() %>%
  
  # show y-value, rounded
  add_tooltip(function(data) round(data$ESR1,1))
  
```

### 4. [{dygraphs}](https://rstudio.github.io/dygraphs/index.html)  

#### Interactive plots for time series data. It is an R interface to the 'dygraphs' JavaScript charting library.  

#### 4.1 Time series plot  
```{r }
# prepare data
data(AirPassengers)
d <- AirPassengers

# make plot
dygraph(d, main = "Airline Passenger Numbers 1949-1960", ylab = "Passengers (Thousands)",
        width = 700, height = 500) %>%
  
  dyOptions(colors = rainbow_hcl(5)[3]) %>%
  
  dySeries(drawPoints = TRUE, pointSize = 1.5) %>%
  
  dyOptions(fillGraph = TRUE, fillAlpha = 0.2, drawGrid = FALSE) %>%
  
  dyRangeSelector()
  
```

### 5. [{plotly}](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf)   

#### Interactive web graphics from R via plotly's JavaScript graphing library https://plot.ly/r/. 


#### 5.1 Boxplot    

``` {r fig.width = 2, fig.height = 1}
plot_ly(mData, y = ~GEX, color = ~geneCluster, type = 'box' )
```

#### 5.2 Correlation heatmap  

``` {r fig.width = 2, fig.height = 1.7}
d = NeveG200_data[1:10, 1:10]
cormat = cor(d)
heatmaply(cormat)
```

#### 5.3 Transform ggplot2 figure to plotly 

``` {r fig.width = 2, fig.height = 1.7}
fig <- qplot(data = Data, ageY, MEGF9, geom=c('point','smooth')) 
ggplotly(fig)
```


#### 5.4 Plot and annotate map (R ?plot_ly example) 

world{maps}, canadian.cities{maps) datasets

``` {r fig.width = 3, fig.height = 2}
ggplot2::map_data("world", "canada") %>%
  group_by(group) %>%
  plot_ly(x = ~long, y = ~lat) %>%
  add_polygons(hoverinfo = "none") %>%
  add_markers(text = ~paste(name, "<br />", pop), hoverinfo = "text",
              data = maps::canada.cities) %>%
  layout(showlegend = FALSE)
```


#### 5.5 Surface plot (R ?plot_ly example)  

volcano{datasets}: Topographic Information on Auckland's Maunga Whau Volcano

```{r fig.width = 3, fig.height = 2}
x <- seq_len(nrow(volcano)) + 100
y <- seq_len(ncol(volcano)) + 500
plot_ly() %>% add_surface(x = ~x, y = ~y, z = ~volcano)
```

