---
title: "R Graphics Devices"
output: word_document
---

```{r setup, cache=FALSE, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width=4, fig.height=4, dpi=300, warning=FALSE, message=FALSE)
```
``` {r include = FALSE }
oldPar = par()
```

## I. Static Displays

A device in R is the medium to produce a graphical output. This can be the *screen* device or a *file* device.    
The command ?Devices lists all available graphics devices.

##### --------------------------------------------------------------------------------------------------------

### 1. Set Environment  

``` {r  warning = FALSE}
setwd("/Users/balaji/Downloads/plots")
load('/Users/balaji/Downloads/plots/helperFunctions.RData')
load('/Users/balaji/Downloads/plots/NeveG.RData')
Data <- NeveG_data[, -(1:4)]
Info <- NeveG_data[,1:4]

library(datasets); library(forecast);library(ggfortify);
library(grid); library(gridExtra); library(OpenStreetMap); library(rgdal)
```

### 2. The *screen* device 

Commonly the default device, which can be launched with a plotting function or command windows(), quartz(), or x11()on Windows, Mac, and Unix/Linux, respectively.  

Multiple devices can be opened, but plotting can be directed only to one device at a time; every open device is assigned an integer.  
-  dev.list() lists all open devices  
-  dev.cur() lists active device  
-  dev.set(integer) activates device  
-  dev.off() closes device  

- 2.1 Open and navigate multiple devices  

``` {r results = 'hide', fig.show = 'hide'}
dev.list()

plot(1:10)
  dev.cur()
    dev.list()
    
windows(w=4,h=3)
  pie(1:10)

windows(w=4,h=3)
  barplot(rep(1,10))

windows(w=4,h=3)
  plotPoints()
    
dev.list()
dev.cur()

dev.off(3)

dev.set(4)
title('windows 4')
dev.off(4)

dev.list()

dev.off(5)

dev.off()

```


### 3. Examples of graphics *file* devices 

- 3.1 PDF (vector format)  

Good for printing (good quality of line-type graphics), however inefficient if a plot has a lot of data points.

``` {r message = FALSE, fig.width = 10, fig.height = 10}

# open graphics device
#pdf('Fig_PairsHist.pdf', width = 6, height = 6)

# make figure (page 1)
par(oma = c(2,2,2,2))
    pairs(Data[,1:4], 
          lower.panel = NULL,
          upper.panel = function(x,y, ...){
            points(x, y, col = 'darkgrey', pch=16)
              abline(lsfit(x,y), col = 'red')})
    mtext('Fig_PairsHist.pdf, Page 1', outer=T, adj = 0, side =1)
 # make figure (page 2)         
par(mfrow = c(3,3))
    for (i in 1:ncol(Data)){
      hist(Data[,i], col = 'gold', 
           main = colnames(Data)[i], xlab = 'log2 GEX')
    }
mtext('Fig_PairsHist.pdf, Page 2', outer=T, adj= 0, side = 1 )

# close device    
#dev.off()        
```

- 3.2 TIFF/PNG (bitmap format)    

Good for plots with large number of points, web-based plots, or journal publishing.

``` {r fig.width = 8, fig.height = 6, dpi = 300}

## PLOT TIFF

#.. prepare data
data(AirPassengers)
d <- AirPassengers
d.arima <- auto.arima(d)
d.forecast <- forecast(d.arima, level = c(99), h = 48)

#.. open graphics device
#tiff('Fig_TimeSeries.tiff', width = 5, height = 4, units = 'in', res = 300, compress='lzw')

#.. make figure
p1 <- autoplot(d.forecast, conf.int.fill = 'yellow',  conf.int.alpha = 0.5, ts.col = 'blue3', predict.col = 'red') + theme_bw()
p2 <-  labs(x = "Year", y = "Passengers (Thousands)", title = "Airline Passenger Numbers 1949-1960")
grid.arrange(p1+p2, right = textGrob("ARIMA 99CI for the forecast", rot = -90, vjust = 1))

#.. close device    
#dev.off()
```


``` {r fig.width = 3, fig.height = 3, dpi = 150}
## PLOT PNGs

#..prepare data

#..get coordinates
#launchMapHelper()

map.type = c("bing", "apple-iphoto", "stamen-watercolor")
UL <- c(lat= 39.06184913429154,   lon= -78.7060546875)
LR <- c(lat= 35.24561909420681,   lon= -74.1796875)
text.colo <- c('white', 'black', 'black')

#.. open graphics device
#png('Map_%1d.png', units='in', width = 4, height = 4, res = 150)

#..make figures
for (i in 1:length(map.type)){
  map <- openmap(UL, LR, minNumTiles=10, type= map.type[i])
    plot(map)
      mtext(map.type[i], side = 1, adj=0.8, line = 2, col = text.colo[i])
}

#.. close device
#dev.off()
```
- 3.3 Exercise: output a figure with:
1. File 1 pdf(): page size in inches: 10w x 8h; 2-pages in one file;
2. File 2 tiff(): page size in inches: 44w x 4h; res=300; compressed (lzw);
3. use R pre-loaded data mtcars and plots of your choice

```{r fig.width=10, fig.height=10}
data(mtcars)
head(mtcars)
?mtcars5
brand = sapply(strsplit(rownames(mtcars), ' '), function(x) x[1])
table(brand)

#pdf('File_1.pdf', width=10, height=8)
ord = order(mtcars$hp, decreasing = TRUE)
plot(mtcars$hp[ord], ylab='gross horsepower', xlab='rank',
     main="Data obtained in mid-1970s", type='h', lwd=1, ylim=c(0,max(mtcars$hp)))
text((1:nrow(mtcars)+0.3), mtcars$hp[ord], rownames(mtcars)[ord],
     srt=90, cex=1, adj=1, col='blue')

par(oma=c(0,8,0,2))
dat = mtcars[,c('cyl','carb')]
barplot(t(dat[rev(ord),]), beside=F, horiz = T, las=2, xlab='count', main='Data ordered by "gross horsepower"', col=c('grey40','grey90'))
legend('bottomright', c('cylinders','carburetors'), fill=c('grey40','grey90'))
#dev.off()

#tiff('File_2.tiff', units='in', res=300, width=4, height=4, compression = 'lzw')
par(pty='s')
plot(mtcars$carb, mtcars$qsec, cex=0.5, pch=16, main='Data otained in mid-1970s')
abline(lm(mtcars$qsec~mtcars$carb))
#dev.off()


```


- [10 tips for making R graphics look good]:(http://blog.revolutionanalytics.com/2009/01/10-tips-for-making-your-r-graphics-look-their-best.html)


