---
title: "Network plots - igraph"
output:
  word_document: default
  html_document: default
---

```{r setup, cache=FALSE, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_chunk$set(fig.width=4, fig.height=4, dpi=300, warning=FALSE, message=FALSE)
```
``` {r include = FALSE }
oldPar = par()
```

## I. Package igraph
D or U, for a directed or undirected graph
N for a named graph (where nodes have a name attribute)
W for a weighted graph (where edges have a weight attribute)
B for a bipartite (two-mode) graph (where nodes have a type attribute)


##### --------------------------------------------------------------------------------------------------------

### 1. Set Environment  

``` {r  warning = FALSE}
setwd("/Users/balaji/Downloads/plots")

library(igraph)

```

### 2. Create graph  

- 2.1 Input: edge description  

"-" undirected ties
"+-" or "-+"  directed ties pointing left & right  
"++"  symmetric ties  
":" sets of vertices

``` {r fig.width = 12, fig.height = 6}

# small undirected network
g1  <- graph_from_literal(a---b, b---c, d---a, e---a) # the number of dashes doesn't matter
g2 <- graph_from_literal(a+--b, b--+c, d:e+++a) # the number of dashes doesn't matter
 
# plot network
par(mfrow = c(1,2))

plot(g1, main = 'Undirected network')
plot(g2, , main = 'Directed network')
    
# print adjacency matrix
g1[]
g2[]

# print object description
print(g1)
print(g2)

# list graph nodes (vertices)
V(g1)

# list graph edges (connections)
E(g2)

# access attributes
V(g1)$name

# add/change attributes

#..node degree (number of nodes connected to a given node)
nodeDegree <- degree(g1)
V(g1)$degree <- nodeDegree

nodeDegree <- degree(g2)
V(g2)$degree <- nodeDegree

#.. change node names
oldNames <- V(g2)$name
V(g2)$name <- toupper(oldNames)

#.. add edge weight
E(g1)$weight <- c(5,3,3,1)
E(g2)$weight <- c(5,3,3,1,4,2)

# list all atributes

#.. vertices
vertex_attr(g1)
vertex_attr(g2, 'degree')

#.. edges
edge_attr(g1)
edge_attr(g2, 'weight')

```

- 2.1 Input: build-in graphs  

``` {r fig.width = 12, fig.height = 12}

par(mfrow = c(2,4))

# empty graph
g <- make_empty_graph(10)
plot(g, main = 'Empty graph')

# full graph
g <- make_full_graph(10)
plot(g, main = 'Full graph', vertex.label = NA)

# star graph
g <- make_star(10)
plot(g, main = 'Star graph', vertex.label = NA)

# tree graph
g <- make_tree(40, children =3)
plot(g, main = 'Tree graph', vertex.label = NA)

# ring graph
g <- make_ring(10)
plot(g, main = 'Ring graph', vertex.label = NA)

# Erdos-Renyi random graph model
#..n is number of nodes, m is the number of edges
g <- sample_gnm(n=20, m=10) 
plot(g, main = 'Random\ngraph model', vertex.label = NA)


# Barabasi-Albert preferential attachment model for scale-free graphs
g <-  sample_pa(n=100, power=1, m=1,  directed=F)
plot(g, main = 'Scale-free\ngraph nmodel', vertex.label = NA, vertex.size = 8)
```

- 2.1 Input: data file (edge list or adjacency matrix) 

2.1.1 Undirected graph (gene co-expression network)
``` {r }
setwd("/Users/balaji/Downloads/plots")
# load gene expression data
load('NeveG200.RData')
Data <- NeveG200_data
Info <- NeveG200_info

# generate correlation matrix

cormat_ini = abs(cor(t(Data)))

# create network
cn = graph.adjacency(cormat_ini, mode='undirected', weighted=T, diag=FALSE)
cn

# remove weak connections
cormat <- cormat_ini
cormat[ abs(cormat) < 0.8 ] = 0
cnet = graph.adjacency(cormat, mode='undirected', weighted=T, diag=FALSE)
cnet
plot(cnet, vertex.label = NA, vertex.size = 5, vertex.color = 'lightblue')

deg <- degree(cnet)
V(cnet)$degree <- deg


dd <- degree.distribution(cnet, cumulative=T, mode="all")
plot(dd, pch=19, cex=1, col="orange", xlab="Degree", ylab="Cumulative Frequency",
     main = 'Degree distribution')

# remove unconnected nodes
cnet <- delete.vertices( cnet, which(degree(cnet) <= 0))
plot(cnet, vertex.label = NA, vertex.size = 5, vertex.color = 'lightblue')

# select largest connected sub-graph
comm <- fastgreedy.community(cnet)
cluster.comm <- cluster_fast_greedy(cnet)
plot(cluster.comm, cnet, vertex.size = 6, vertex.label = NA)

# Final network
net <- induced.subgraph(cnet, which(membership(comm) <= 3))
plot(net, vertex.label = NA, vertex.size = 5, vertex.color = 'lightblue')

```

### 3. Customize graph

``` {r}

# set vertex features
plot(net, vertex.label =  NA, vertex.size = V(net)$degree,
     vertex.label.cex= 5, 
     vertex.color = 'skyblue',  vertex.frame.color = 'navyblue')

# set edge features   
plot(net, vertex.label =  NA, vertex.size = V(net)$degree,
      vertex.label.cex= 5, 
      vertex.color = 'skyblue',  vertex.frame.color = 'navyblue',
      edge.color = 'grey', edge.width = E(net)$weight)
        

# set layout
lay = layout.kamada.kawai(net)
plot(net, vertex.label =  NA, vertex.size = V(net)$degree,
    vertex.label.cex= 5, 
    vertex.color = 'skyblue',  vertex.frame.color = 'navyblue',
    edge.color = 'grey', edge.width = E(net)$weight,
    layout = lay)

# set layout
lay = layout.grid(net)
plot(net, vertex.label =  NA, vertex.size = V(net)$degree,
    vertex.label.cex= 5, 
    vertex.color = 'skyblue',  vertex.frame.color = 'navyblue',
    edge.color = 'grey', edge.width = E(net)$weight,
    layout = lay)
  
# set layout
lay = layout.circle(net)
plot(net, vertex.label =  NA, vertex.size = V(net)$degree,
    vertex.label.cex= 5, 
    vertex.color = 'skyblue',  vertex.frame.color = 'navyblue',
    edge.color = 'grey', edge.width = E(net)$weight,
    layout = lay)

# move node manually
lay[1,] = lay[1,] + c(1,2)
plot(net, vertex.label =  NA, vertex.size = V(net)$degree,
    vertex.label.cex= 5, 
    vertex.color = 'skyblue',  vertex.frame.color = 'navyblue',
    edge.color = 'grey', edge.width = E(net)$weight,
    layout = lay)


# default layout + add.names
set.seed(100)
lay = layout.auto
plot(net, vertex.label = V(net)$name, vertex.label.cex = 0.4,
    vertex.size = V(net)$degree,
    vertex.label.cex= 5, 
    vertex.color = 'skyblue',  vertex.frame.color = 'navyblue',
    edge.color = 'grey', edge.width = E(net)$weight,
    layout = lay)



```


### Network of R packages 

``` {r fig.width = 14, fig.height = 14}

#install.packages('miniCRAN')
library(miniCRAN)

lib1 <- 'gplots'
graph1 <- makeDepGraph(lib1, enhances=TRUE)
set.seed(10)
plot(graph1, legendPosition = c(-1, 1), vertex.size=20)

lib2 <- 'igraph'
graph2 <- makeDepGraph(lib2, enhances=TRUE)
set.seed(10)
plot(graph2, legendPosition = c(-1, 1), vertex.size=20)

lib3 <- 'ggplot2'
graph3 <- makeDepGraph(lib3, enhances=TRUE)
set.seed(10)
plot(graph3, legendPosition = c(-1, 1), vertex.size=20)

graph <- makeDepGraph(c(lib1, lib2, lib3), enhances=TRUE)
set.seed(10)
plot(graph, legendPosition = c(-1, 1), vertex.size=20)


